Properties of Winning Iterated Prisoner's Dilemma Strategies.

### 2025-07-28 Reply/Point-by-Point-Reply.md

# Point-by-Point Reply

First, we would like to thank the editor and reviewers for their valuable
feedback. Their comments have been helpful and constructive, and we have
revised our manuscript accordingly. In particular, we have made the following
changes:

- Motivated by a comment from Reviewer #1, we have expanded the discussion to
  include further work and revised the introduction to address why such a study
  was not feasible until now.

- Reviewer #1 suggested adding more detail about the Axelrod-Python packages. We
  have included additional explanations to clarify this aspect of our work.

- As suggested by Reviewer #1, we have added references to relevant statements
  throughout the manuscript.

- In line with Reviewer #1's suggestion, we have enriched the captions to
  provide more details for each figure.

- Reviewer #2's comments, we have clarified the measures in the main text, which
  were previously only explained in the figure captions.

# Editor Remarks to Author:

Comment: Thank you very much for submitting your manuscript "Properties of Winning
Iterated Prisoner's Dilemma Strategies" for consideration at PLOS Computational
Biology. As with all papers reviewed by the journal, your manuscript was
reviewed by members of the editorial board and by several independent reviewers.
The reviewers appreciated the attention to an important topic. Based on the
reviews, we are likely to accept this manuscript for publication, providing that
you modify the manuscript according to the review recommendations.

Reply: Thank you for considering our paper and for providing positive feedback.
The reviewers’ comments have been very helpful.  Please find all details aboutx
how we addressed each comment raised by the reiweres in our response to the
reviewers below.

# Reviewer 1

Comment: Summary: This paper explores the performance of strategies in the IPD.
This research analyses 195 strategies across thousands of computer tournaments,
examining what makes certain strategies more successful in diverse IPD
environments. The study used four different types of tournaments: Standard
tournaments, Noisy tournaments, Probabilistic ending tournaments and Noisy
probabilistic ending tournaments. They find that no single strategy excels in
all settings. However, several key traits appear in successful strategies,
particularly the ability to adjust to the environment and population dynamics.

Thanks for this submission, I find the text easy to read and all the concepts
well explained.

Reply: Thank you for carefully reading our manuscript, and for the constructive
feedback below.

---

Comment: This seems like a low-hanging fruit idea for someone like
Axelrod-Python. If they have all of these strategies, what is the best, after
all, they maintain and add new strategies to the library, why do you think this
hasn't been done before? I feel the data analysis and the interpretation is
sound, but maybe can you add other efforts to do the same, and what do you do
differently?

Reply: We agree with the reviewer’s comment that this could be considered a
``low-hanging fruit'' idea for the developers of the Axelrod-Python library. We
understand that the reviewer is asking whether this type of analysis has been
done before; if so, what did we do differently?

To start, we would like to emphasize that this analysis would not have been
feasible before the development of the Axelrod-Python library. Previously,
researchers would have needed to locate, understand, and implement code for each
strategy from the literature in order to conduct a large-scale tournament.
Additionally, they would have had to develop all the necessary code to manage
the tournament, a challenging task for any researcher. The creation of the
Axelrod-Python package has made such an analysis feasible. This package is the
result of a collaborative effort that has produced a substantial database of
strategies from the literature. We would also like to highlight that the
library’s open-source nature has encouraged contributions from a broad user
base.

Since the package’s development, we believe the reasons for limited
comprehensive analyses are rooted in field conventions. In our field,
tournaments or evolutionary dynamics often rely on a select list of hand picked
strategies chosen by modelers, typically based on specific properties they wish
to examine. This practice may stem from certain misconceptions, such as the
assumption that because Tit-for-Tat performs relatively well, testing against
only Tit-for-Tat variants is sufficient. Another misconception may arise from
the Press & Dyson result, which suggests that lower memory strategies will
always dominate in pairwise interactions, leading some researchers to consider
only memory-one strategies. However, this result strictly applies only to
pairwise interactions.

We did find one example where the setup of the tournament itself was questioned.
Specifically, in Is Tit-for-Tat the Answer? On the Conclusions Drawn from
Axelrod's Tournaments (Rapoport, 2015), Rapoport examined the effects of
variations in format, objective criteria, and payoff values on tournament
outcomes. However, their study analyzed different effects than ours and used a
single tournament setup with a specific selection of strategies. They focused on
Axelrod’s original tournament and demonstrated how certain changes affected the
outcome. For example, including strategies like the “kingmakers” positively
impacted the performance of Tit For Tat.

Changes: We have updated the introduction to clearly state why such an analysis
has not been done before, and we discuss the work of Rapoport (2015).

---

Comment: You mention somewhere that there are evolutionary approaches to
this, specifically evolutionary game theory approaches. Maybe adding why you
chose to do computer simulations instead? (I know that EGT has some constraints,
but highlight why you do how you do it to overcome those constraints).

Reply: Excellent observation. Many strategies, such as Win-Stay-Lose-Shift
and Generous Tit For Tat, emerged due to their strong performance in
evolutionary dynamics. Axelrod's original work involved computer tournaments, so
we chose to remain consistent with this approach since this comprehensive study
had not yet been undertaken.

Moreover, in evolutionary dynamics, there are several factors to consider, such
as which evolutionary process to apply and the high computational cost. However,
it remains an interesting question. For example, we could consider strategy
performance in terms of fixation probabilities in the case of low mutation
rates. This is future work we would be eager to pursue.

Changes: We've added a paragraph to the discussion section outlining
potential extensions into evolutionary dynamics.

---

Comment: I don't know if it's enough to say that you used Axelrod-Python,
but maybe also add how these tournaments are performed? I haven't used the tool,
so I don't know how they simulate those exactly. I feel the library itself is
central to how the results are generated, so maybe expand a bit on that.

Reply: Thank you for this suggestion. We agree that many readers may not be
familiar with the Axelrod-Python package.

Changes: We've expanded the methodology section to include more details on how
to run tournaments using the Axelrod library. We have included a figure showing
a code example of how to run such a tournament.

---

Comment: Check for references when you make statements.

Reply: We thank the reviewer for this comment.

Changes: We have added references to the sentences indicated by the reviewer
and in addition the following sentences:

- "Speciﬁcally, much like the recognized value of diversity in training datasets, such as variations in image perspective,..."

- "Already well-known in the AI/ML literature, adding noise to training data
  leads to more robust models"

---

Comment: I believe it's a matter of style, but I prefer if figures are
self-explanatory. As you did with Figure 1, where you guide the reader through
what is shown, Table 4 and Figures 3,4 and 5 lack some context in the figure
labels, especially in figures like 5, some guidance is appreciated.

Reply: We agree that figures should be self-explanatory.

Changes: We have included additional details in the figures and tables that were
previously missing. These were Table 4 and Figures 3, 4, and 5 (now Figures 4,
5, and 6).

---

# Reviewer 2

Comment: The authors uncover properties of winning strategies in the iterated
Prisoner’s Dilemma (IPD), one of the most renowned paradigms for studying
cooperation over the past decades. They consider a large collection of 195
strategies in thousands of computer tournaments and conduct a thorough analysis
of their performance. Their conclusions refine the properties described by
Axelrod: a successful strategy needs to be nice, provocable, generous, somewhat
envious, clever, and adaptable to the environment.

There are several aspects of this work that I appreciate greatly. First, the
inclusiveness of the study design is commendable: it not only includes all kinds
of strategies from IPD literature but also incorporates variations in the
tournaments (such as random noise and match length). Second, their findings
challenge some of Axelrod’s suggestions, particularly the advice to “not be
clever” and “not be envious.” Additionally, I find the property of being
adaptable to the environment—explained by the authors as a strategy’s
cooperation rate compared to the average cooperation rate in a tournament—very
inspiring.

Reply: Thank you for your comprehensive summary of our work! We appreciate the
comments.

---

Comment: On page 8, I suggest including the explanations for C_max,
C_median, C_mean, and SSE error in the main text, as they currently seem to be
defined only in figures and tables.

Reply: Thank you for this suggestion. These measures should indeed be
explained in the main text as well.

Changes: We have included explanations of each measure in the main text, in
section "Model".

---

Comment:  In Table 3 on page 9, there is a contradiction between "A
strategy's cooperating rate divided by the minimum" and "C_min/C_r."

Reply: Correct. That is a typo. We have addressed this.

---

Comment: On page 13, the sentence "envious strategies capable of both
exploiting and not their opponents…" appears to be incomplete.

Reply: This is indeed correct; the sentence is incomplete. The sentence should
read: ``We also showed that while the type of exploitation attempted by ZDs is
not typically effective in standard tournaments, envious strategies capable of
both exploiting and not their opponents can be highly successful''.

### 2025-07-28 paper/main.tex
Submission cb (#38)
* drafts sections Data Collection and Top Ranked Strategies

* working on main paper for the special issue

* finish off top performance analysis for paper and si

* finish SI section analysis with all data and clean notebooks

The notebooks now include code to run the analysis for both the
main paper and the SI

* last corrections to the paper and commit si

* commit new notebooks

* Some copy edits for special issue version (#34)

* Update main.tex

* Update regression_table.tex

Fix Typo

* change NG affiliation, note that VK and MH contributed equally + ref style

* add

In this paper, we evaluate the performance of a significant number of \IPD{}
strategies across a diverse array of tournaments. Many of the strategies used in
our analysis are drawn from well-known and named strategies in \IPD{}
literature, including previous tournament winners. This contrasts with other
work that may have randomly generated essentially arbitrary strategies, often
constrained to specific classes such as memory-one strategies or those of a
certain structural form like finite state machines or deterministic memory-two
strategies. Furthermore, our tournaments encompass variations, including standard
tournaments resembling Axelrod's original ones, tournaments with noise,
probabilistic match length, and both noise and probabilistic match length. This
diversity in strategies and tournament types provides new insights and tests
earlier claims in alternative settings against known powerful strategies.
More specifically, we show that the previous tournament
winners are lacking against large enough opponent pools; they do not appear
among the top-performing strategies anymore. This could be due to likely
suffering from a lack of diversity in the strategies they were trained/tested
against, finding it hard to adapt to the new strategies.

@@

Similarly, our results could suggest an explanation regarding the intuitively
unexpected effectiveness of memory-one strategies historically. Given that among
the important features associated with success are the relative cooperation rate
to the population average and the four memory-one probabilities of cooperating
conditional on the previous round of play, these features can be optimized by a
memory-one strategy such as \TFT. Usage of more history becomes valuable when
there are exploitable opponent patterns. This is indicated by the importance of
SSE as a feature, showing that the first-approximation provided by a memory-one
strategy is no longer sufficient.
These results highlight a central idea in evolutionary game theory in this
context: the fitness landscape is a function of the population (where fitness in
this case is tournament performance). While that may seem obvious now, it shows
why historical tournament results on small or arbitrary populations of
strategies have so often failed to produce generalizable results.

@@

The raw and processed datasets have been made publicly available~\cite{data,
raw_data} and can be used for further analysis and insights.

>>>>

In this paper, we evaluate the performance of a significant number of \IPD{}
strategies across a diverse array of tournaments. Many of the strategies used in
our analysis are drawn from well-known and named strategies in \IPD{}
literature, including previous tournament winners.
This contrasts with other work that is often constrained to specific classes such as 
as memory-one strategies or those of a
certain structural form like finite state machines or deterministic memory-two
strategies. Furthermore, our tournaments encompass variations, including standard
tournaments resembling Axelrod's original ones, tournaments with noise,
probabilistic match length, and both noise and probabilistic match length. This
diversity in strategies and tournament types provides new insights and tests
earlier claims in alternative settings against known powerful strategies.
More specifically, we show that the previous tournament
winners are lacking against large enough opponent pools; they do not appear
among the top-performing strategies anymore. This could be due to likely
suffering from a lack of diversity in the strategies they were trained/tested
against, finding it hard to adapt to the new strategies.

@@

Our results may also help explain the historically unexpected effectiveness of
memory-one strategies~\cite{Camerer2011}. The success of these strategies
contradicts the intuitive assumption that a longer memory and therefore more
information would yield better strategic performance~\cite{Binmore1987}. Given
that among the important features associated with success are the relative
cooperation rate to the population average and the four memory-one probabilities
of cooperating conditional on the previous round of play, these features can be
optimized by a memory-one strategy such as \TFT. Usage of more history becomes
valuable when there are exploitable opponent patterns. This is indicated by the
importance of SSE as a feature, showing that the first-approximation provided by
a memory-one strategy is no longer sufficient. These results highlight a central
idea in evolutionary game theory in this context: the fitness landscape is a
function of the population (where fitness in this case is tournament
performance)~\cite{Hofbauer2003}. While that may seem obvious now, it shows why
historical tournament results on small or arbitrary populations of strategies
have so often failed to produce generalizable results.

@@

The raw and processed datasets have been made publicly available~\cite{raw_data, data}
and can be used for further analysis and insights.

### 2025-07-28 paper/regression_on_median_score.tex
Submission cb (#38)
* drafts sections Data Collection and Top Ranked Strategies

* working on main paper for the special issue

* finish off top performance analysis for paper and si

* finish SI section analysis with all data and clean notebooks

The notebooks now include code to run the analysis for both the
main paper and the SI

* last corrections to the paper and commit si

* commit new notebooks

* Some copy edits for special issue version (#34)

* Update main.tex

* Update regression_table.tex

Fix Typo

* change NG affiliation, note that VK and MH contributed equally + ref style

* add

A multivariate linear regression has also been fitted to model the relationship between
the features and the median score. The features included are given
by Table~\ref{table:linear_regression_on_median} alongside their corresponding \(p\) values
in the distinct tournaments and their regression coefficients.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
\resizebox{\textwidth}{!}{
\begin{tabular}{lggccggccgg}
    \toprule
    & \multicolumn{2}{g}{Standard} & \multicolumn{2}{c}{Noisy} & \multicolumn{2}{g}{Probabilistic ending} & \multicolumn{2}{c}{Noisy probabilistic ending} & \multicolumn{2}{g}{Overall}\\
    \midrule
    & \multicolumn{2}{g}{\(R\) adjusted: 0.576} & \multicolumn{2}{c}{\(R\) adjusted: 0.679} & \multicolumn{2}{g}{\(R\) adjusted: 0.816} & \multicolumn{2}{c}{\(R\) adjusted: 0.930} & \multicolumn{2}{g}{\(R\) adjusted: 0.318} \\
    {} &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value \\
    \midrule
    $CC$ to $C$ rate   &  0.043 &  0.000 & -0.380 &  0.000 &  0.224 &  0.000 &  0.078 &    0.0 &  0.308 &    0.0 \\
    $CD$ to $C$ rate   & -0.325 &  0.000 &  0.124 &  0.000 &  0.060 &  0.000 &  0.073 &    0.0 & -0.014 &    0.0 \\
    $C_r$ / $C_{max}$  &      - &      - & -1.044 &  0.000 &      - &      - & -1.251 &    0.0 &      - &      - \\
    $C_r$ / $C_{mean}$ &  0.553 &  0.000 & -0.101 &  0.000 & -1.136 &  0.000 & -0.089 &    0.0 & -0.665 &    0.0 \\
    $C_{max}$          &  0.059 &  0.000 &      - &      - & -0.044 &  0.086 & -1.396 &    0.0 &      - &      - \\
    $C_{mean}$         &  1.837 &  0.000 &  3.078 &  0.000 &  1.506 &  0.000 &  3.645 &    0.0 &      - &      - \\
    $C_{min}$          &  0.156 &  0.000 &  1.528 &  0.000 &  0.311 &  0.000 &      - &      - &      - &      - \\
    $C_{min}$ / $C_r$  & -0.049 &  0.000 & -0.378 &  0.000 & -0.204 &  0.000 &      - &      - & -0.257 &    0.0 \\
    $DC$ to $C$ rate   & -0.204 &  0.000 &  0.074 &  0.000 &  0.066 &  0.000 &  0.066 &    0.0 &  0.007 &    0.0 \\
    $k$                & -0.000 &  0.853 & -0.000 &  0.987 &  0.000 &  0.008 &  0.000 &    0.0 &      - &      - \\
    $n$                & -0.000 &  0.000 &      - &      - &      - &      - &      - &      - &      - &      - \\
    $p_e$              &      - &      - &      - &      - &  0.025 &  0.000 & -0.095 &    0.0 &      - &      - \\
    $p_n$              &      - &      - &  0.124 &  0.000 &      - &      - &      - &      - &      - &      - \\
    SSE                & -0.294 &  0.000 & -0.319 &  0.000 &  0.055 &  0.000 &  0.010 &    0.0 & -0.015 &    0.0 \\
    constant           &  0.925 &  0.000 &  1.536 &  0.000 &  2.466 &  0.000 &  2.299 &    0.0 &  2.924 &    0.0 \\
    memory usage       &  0.010 &  0.000 & -0.004 &  0.000 &      - &      - &      - &      - &      - &      - \\
    \bottomrule
\end{tabular}}
\end{center}
\caption{Results of multivariate linear regressions with the median score as the dependent variable.
\(R\) squared is reported for each model.}
\label{table:linear_regression_on_median}
\end{table}

>>>>

A multivariate linear regression has also been fitted to model the relationship
between the features and the median score. The features included are given in
Table~\ref{table:linear_regression_on_median} alongside their corresponding
\(p\) values in distinct tournaments and their regression coefficients.
Table~\ref{table:linear_regression_on_median} provides the results when
considering tournaments with \(p_n \leq 0.1\) and \(p_e \leq 0.1\). The results
when the probabilities are not constrained are given in
Table~\ref{table:linear_regression_on_median_full_dataset}.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
\resizebox{.8\textwidth}{!}{
\begin{tabular}{lggccggccgg}
    \toprule
    & \multicolumn{2}{g}{Standard} & \multicolumn{2}{c}{Noisy} & \multicolumn{2}{g}{Probabilistic ending} & \multicolumn{2}{c}{Noisy probabilistic ending}\\
    \midrule
    & \multicolumn{2}{g}{\(R\) adjusted: 0.576} & \multicolumn{2}{c}{\(R\) adjusted: 0.561} & \multicolumn{2}{g}{\(R\) adjusted: 0.488} & \multicolumn{2}{c}{\(R\) adjusted: 0.762}\\
    {} &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value \\
    \midrule
    constant             &  0.928 &  0.000 &  1.082 &  0.000 &  1.259 &  0.000 &  1.642 &  0.000 \\
    $CC$ to $C$ rate     &  0.043 &  0.000 &  0.104 &  0.000 &  0.024 &  0.000 & -0.032 &  0.000 \\
    $CD$ to $C$ rate     & -0.325 &  0.000 & -0.052 &  0.000 & -0.229 &  0.000 & -0.110 &  0.000 \\
    $DC$ to $C$ rate     & -0.204 &  0.000 & -0.076 &  0.000 & -0.102 &  0.000 & -0.070 &  0.000 \\
    SSE                  & -0.294 &  0.000 & -0.186 &  0.000 & -0.131 &  0.000 & -0.109 &  0.000 \\
    $C_{max}$            &  0.056 &  0.000 & -0.060 &  0.011 & -0.005 &  0.849 & -0.240 &  0.000 \\
    $C_{min}$            &  0.156 &  0.000 & -0.159 &  0.000 &  0.012 &  0.385 &  0.083 &  0.005 \\
    $C_{mean}$           &  1.838 &  0.000 &  2.247 &  0.000 &  1.822 &  0.000 &  2.030 &  0.000 \\
    $C_{min}$ / $C_r$    & -0.049 &  0.000 &  0.040 &  0.000 & -0.018 &  0.000 & -0.065 &  0.000 \\
    $C_r$ / $C_{mean}$   &  0.552 &  0.000 & -0.227 &  0.000 &  0.027 &  0.125 & -0.043 &  0.000 \\
    $k$                  & -0.000 &  0.856 & -0.000 &  0.245 & -0.000 &  0.610 & -0.000 &  0.572 \\
    $n$                  & -0.000 &  0.000 &      - &      - &      - &      - &      - &      - \\
    memory usage         &  0.010 &  0.000 &  0.004 &  0.000 &      - &      - &      - &      - \\
    $C_r$ / $C_{median}$ &      - &      - &  0.252 &  0.000 &  0.181 &  0.000 &      - &      - \\
    $p_n$                &      - &      - &  0.770 &  0.000 &      - &      - &  0.494 &  0.000 \\
    $p_e$                &      - &      - &      - &      - &  0.763 &  0.000 &  0.285 &  0.000 \\
    \bottomrule
\end{tabular}}
\end{center}
\caption{Results of multivariate linear regressions with the median score as the dependent variable.
\(R\) squared is reported for each model. For noisy tournaments $p_n \leq 0.1$
and for probabilistic ending tournaments $p_e \leq 0.1$.}
\label{table:linear_regression_on_median}
\end{table}

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
\resizebox{.8\textwidth}{!}{
\begin{tabular}{lggccggccgg}
    \toprule
    & \multicolumn{2}{g}{Standard} & \multicolumn{2}{c}{Noisy} & \multicolumn{2}{g}{Probabilistic ending} & \multicolumn{2}{c}{Noisy probabilistic ending}\\
    \midrule
    & \multicolumn{2}{g}{\(R\) adjusted: 0.575} & \multicolumn{2}{c}{\(R\) adjusted: 0.666} & \multicolumn{2}{g}{\(R\) adjusted: 0.816} & \multicolumn{2}{c}{\(R\) adjusted: 0.869} \\
    {} &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value &  Coefficient &  \(p\)-value\\
    \midrule
    constant           &  0.928 &  0.000 &  2.143 &  0.000 &  2.466 &  0.000 &  1.824 &  0.000 \\
    $CC$ to $C$ rate   &  0.043 &  0.000 & -0.468 &  0.000 &  0.223 &  0.000 & -0.008 &  0.000 \\
    $CD$ to $C$ rate   & -0.325 &  0.000 &  0.105 &  0.000 &  0.060 &  0.000 &  0.074 &  0.000 \\
    $DC$ to $C$ rate   & -0.204 &  0.000 &  0.060 &  0.000 &  0.066 &  0.000 & -0.002 &  0.011 \\
    SSE                & -0.294 &  0.000 & -0.365 &  0.000 &  0.055 &  0.000 & -0.035 &  0.000 \\
    $C_{max}$          &  0.056 &  0.000 &      - &      - & -0.045 &  0.081 & -0.181 &  0.000 \\
    $C_{min}$          &  0.156 &  0.000 &  0.264 &  0.000 &  0.311 &  0.000 &      - &      - \\
    $C_{mean}$         &  1.838 &  0.000 &  2.046 &  0.000 &  1.506 &  0.000 &  2.273 &  0.000 \\
    $C_{min}$ / $C_r$  & -0.049 &  0.000 & -0.252 &  0.000 & -0.204 &  0.000 &      - &      - \\
    $C_r$ / $C_{mean}$ &  0.552 &  0.000 & -0.579 &  0.000 & -1.137 &  0.000 & -0.610 &  0.000 \\
    $k$                & -0.000 &  0.856 &  0.000 &  0.788 &  0.000 &  0.009 &  0.000 &  0.005 \\
    $n$                & -0.000 &  0.000 &      - &      - &      - &      - &      - &      - \\
    memory usage       &  0.010 &  0.000 & -0.006 &  0.000 &      - &      - &      - &      - \\
    $p_n$              &      - &      - &  0.119 &  0.000 &      - &      - &      - &      - \\
    $p_e$              &      - &      - &      - &      - &  0.025 &  0.000 & -0.028 &  0.000 \\
    \bottomrule
\end{tabular}}
\end{center}
\caption{Results of multivariate linear regressions with the median score as the dependent variable.
\(R\) squared is reported for each model.}
\label{table:linear_regression_on_median_full_dataset}
\end{table}
